Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-05-11 paper/manuscript.Rmd
edits minor

The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
observed an unmarked adult male caribou swim between two small islands (Figure
1, see Figure 2 for location), a distance of at least 470 m which took ~9
minutes (~52 m per minute). This observation prompted us to investigate whether
swimming behaviour was common among GPS radio-collared caribou in the population
[n = 29, for collaring details see, @Peignier2019]. We identified swimming
events as two consecutive GPS locations (2-hour relocation rates)  from an
individual occurred on different islands. Northeastern Newfoundland typically
experiences pack ice during winter and caribou are known locally to travel
between islands by walking over the ice. We restricted our GPS data to the
ice-free period of the year (April 1 to December 31) and identified `r nevents`
swimming events over three years (Figure 2). In addition to our own visual and
remotely sensed observations, residents of Fogo Island have also reported
observing caribou swimming between islands on numerous occasions.

In total, 12 of 29 collared female caribou swam between islands in the Fogo
Island archipelago. On average, caribou swam approximately `r neventidMean`
(range = `r neventidMin`-`r neventidMax`) times per year and remained on each
island for approximately `r dur` (range = `r durMin`-`r durMax`) days before
swimming again. We can separate the population into three apparent groups of
individuals: those that swam regularly between islands (n = 3, every ~`r indMean
days`); those that rarely engaged in swimming events (n = 9, every ~ `r ind2Mean
days`); and those that never swam (n = 17).

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, the energetic expenditure
of swimming for quadrupedal mammals is significantly higher than walking or
running, and drowning is also possible [@Miller1985]. Despite the abundance of
coastal and island caribou herds, only a handful of examples have documented
caribou swimming in the ocean [e.g. @Miller1995; @Miller2002;
@Jeffery2007; @Ricca2012]. Caribou may swim between islands in the ocean for
similar reasons they swim in freshwater [@Leblond2016], mainly concerning
predator avoidance or movement between habitats. We propose that one such
explanation, the forage limitation hypothesis, is the most likely explanation
for oceanic swimming for caribou that live on islands in the Fogo Island
archipelago.

>>>>

The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
observed an unmarked adult male caribou swim between two small islands (Figure
1, see Figure 2 for location), a distance of at least 470 m which took ~9
minutes (~52 m per minute). This observation prompted us to investigate whether
swimming behaviour was common among GPS radio-collared caribou in the population
[n = 29, for collaring details see, @Peignier2019]. We identified swimming
events as two consecutive GPS locations (2-hour relocation rates)  from an
individual that occurred on different islands. Northeastern Newfoundland typically
experiences pack ice during winter and caribou are known locally to travel
between islands by walking over the ice. We restricted our GPS data to the
ice-free period of the year (April 1 to December 31) and identified `r nevents`
swimming events over three years (Figure 2). In addition to our own visual and
remotely sensed observations, residents of Fogo Island have also reported regularly
observing caribou swimming between islands.

In total, 12 of 29 collared female caribou swam between islands in the Fogo
Island archipelago. On average, caribou swam approximately `r neventidMean`
(range = `r neventidMin`-`r neventidMax`) times per year and remained on each
island for approximately `r dur` (range = `r durMin`-`r durMax`) days before
swimming again. Based on swimming data, we separated the population into three
apparent groups of individuals: those that swam regularly between islands (n =
3, every ~`r indMean days`); those that rarely engaged in swimming events (n =
9, every ~ `r ind2Mean days`); and those that never swam (n = 17).

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, that have hollow hair shafts to presumably aid in flotation, the energetic expenditure
of swimming for quadrupedal mammals is significantly higher than walking or
running, and drowning is also possible [@Miller1985]. Despite the abundance of
coastal and island caribou herds, only a handful of examples have documented
caribou swimming in the ocean [e.g. @Miller1995; @Miller2002;
@Jeffery2007; @Ricca2012]. Caribou may swim between islands in the ocean for
similar reasons they swim in freshwater [@Leblond2016], mainly concerning
predator avoidance or movement between habitats. We propose that one such
explanation, the forage limitation hypothesis, is the most likely explanation
for oceanic swimming for caribou that live on islands in the Fogo Island
archipelago.
